Analysis of variability

### 2016-07-01 portfolio/analysis/7-separate-models.Rmd
Rework separate model analysis and notes

---
title: "Dive into separate models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
rm(list = ls())
dm <- readRDS("../data-generated/cfec-annual-for-modeling.rds")
nrow(dm)
```

@@

```{r}
dm <- group_by(dm, p_holder) %>%
  mutate(nyr = length(unique(year))) %>%
  mutate(range_div = diff(range(specDiv))) %>%
  as_data_frame() %>%
  filter(nyr >= 10) %>%
  filter(range_div > 0)
nrow(dm)

@@

```{r}
dmv <- dm %>% group_by(p_holder) %>%
  summarise(cv = sd(revenue)/mean(revenue), log_specDiv = mean(log(specDiv)),
    # summarise(cv = mean(revenue), log_specDiv = mean(log(specDiv)),
    length = mean(length, na.rm = TRUE), days = mean(days, na.rm = TRUE),
    strategy = paste(sort(unique(substr(strategy, 1, 1))), collapse = " ")) %>%
  as_data_frame() %>%
  group_by(strategy) %>%
  mutate(n_strategy = n()) %>%
  filter(n_strategy > 5) %>%
  mutate(mean_log_spec = mean(log_specDiv)) %>%
  as_data_frame() %>%
  mutate(scaled_log_days = scale(log(days + 1)), scaled_log_spec = scale(log_specDiv))
dmv$strategy <- as.factor(dmv$strategy)
```

@@

```{r}
length(unique(dm$p_holder))
nrow(dmv)
```

We'll fit a model with `lmer()`. Random intercepts and random slopes for diversity group by strategy. I've also now added a quadratic term for species diversity.

```{r}
m4 <- lmer(log(cv) ~ scaled_log_spec + scaled_log_days + I(scaled_log_spec^2) +
    (1 + scaled_log_spec+I(scaled_log_spec^2)|strategy), data = dmv)
co <- coef(m4)$strategy
co$strategy <- row.names(co)
names(co) <- c("intercept", "log_diversity", "log_days", "log_diversity_squared", "strategy")
arm::display(m4)
```

Because number of days fished is related to species diversity CV decreases both with increases in species diversity and with increases in number of days fished. Therefore for the purposes of visualization I will predict at a few levels of mean species diversity.

```{r}
newdata <- data.frame(select(dmv, -scaled_log_days), scaled_log_days = 0)
newdata <- dmv %>%
  mutate(diversity_cut = cut(log_specDiv, breaks = c(-1, 0.25, 0.5, 0.75, 1e6))) %>%
  group_by(strategy, diversity_cut) %>%
  mutate(scaled_log_days = mean(scaled_log_days))
```

```{r, fig.width=10, fig.height=9}
dmv$prediction <- predict(m4, newdata = newdata)
dmv %>%
  arrange(strategy, log_specDiv) %>%
  ggplot(aes(scaled_log_spec, log(cv), color = scaled_log_days)) + facet_wrap(~strategy) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
  geom_line(aes(y = prediction), color = "orange", lwd = 1)
```

```{r}
dmv %>%
  arrange(strategy, log_specDiv) %>%
  ggplot(aes(scaled_log_spec, log(cv), color = scaled_log_days, group = strategy)) +
  geom_point(alpha = 0.7) +
  geom_line(aes(y = prediction), color = "orange", lwd = 1)
```

Let's get an idea of what a group level predictors would look like. In other words what the effect is like across strategies.

```{r}
co <- co %>%
  mutate(n_strategy = nchar(gsub(" ", "", strategy))) %>%
  inner_join(select(dmv, strategy, mean_log_spec) %>% unique())
ggplot(co, aes(mean_log_spec, log_diversity, label = strategy)) + geom_text() +
  ylab("diversity slope random effect") +
  geom_smooth()
ggplot(co, aes(mean_log_spec, intercept, label = strategy)) + geom_text() +
  ylab("intercept random effect") +
  geom_smooth()
ggplot(co, aes(mean_log_spec, log_diversity_squared, label = strategy)) + geom_text() +
  ylab("log_diversity_squared random effect") +
  geom_smooth()
```

```{r, fig.width=10, fig.height=6}
strategy_grouped = rbind(cbind(co[grep("M",co$strategy),], group="Groundfish"),
  cbind(co[grep("S",co$strategy),], group="Salmon"),
  cbind(co[grep("B|C",co$strategy),], group="Halibut-Sablefish"),
  cbind(co[grep("K|T|D",co$strategy),], group="Crab"),
  cbind(co[grep("G",co$strategy),], group="Herring"))
ggplot(strategy_grouped, aes(mean_log_spec, intercept, label = strategy)) +
  geom_text() + geom_smooth(se = FALSE) +
  facet_wrap(~group, scales = "free_y")
ggplot(strategy_grouped, aes(mean_log_spec, log_diversity, label = strategy)) +
  geom_text() + geom_smooth(se = FALSE) +
   ylab("diversity slope random effect") +
  facet_wrap(~group, scales = "free_y")
ggplot(strategy_grouped, aes(mean_log_spec, log_diversity_squared, label = strategy)) +
  geom_text() + geom_smooth(se = FALSE) +
   ylab("log_diversity_squared random effect") +
  facet_wrap(~group, scales = "free_y")
```

>>>>

---
title: "Analysis of variability"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, autodep = TRUE, warning = FALSE)
```

@@

```{r load-data}
library(dplyr)
library(ggplot2)
library(lme4)
dm <- readRDS("../data-generated/cfec-annual-for-modeling.rds")
nrow(dm)
```

@@

```{r filter-data}
dm <- group_by(dm, p_holder) %>%
  mutate(nyr = length(unique(year))) %>%
  mutate(range_div = diff(range(specDiv))) %>%
  as_data_frame() %>%
  filter(nyr >= 5) %>%
  filter(range_div > 0)
nrow(dm)

@@

```{r condense-data-by-permit-holder}
dmv <- dm %>% group_by(p_holder) %>%
  summarise(cv = sd(revenue)/mean(revenue), log_specDiv = mean(log(specDiv)),
    # summarise(cv = mean(revenue), log_specDiv = mean(log(specDiv)),
    length = mean(log(length+1), na.rm = TRUE), days = mean(days, na.rm = TRUE),
    strategy = paste(sort(unique(substr(strategy, 1, 1))), collapse = " ")) %>%
  as_data_frame() %>%
  group_by(strategy) %>%
  mutate(n_strategy = n()) %>%
  filter(n_strategy > 4) %>%
  mutate(mean_log_spec = mean(log_specDiv)) %>%
  as_data_frame() %>%
  mutate(scaled_log_days = scale(log(days + 1)), scaled_log_spec = scale(log_specDiv),
    scaled_log_length = scale(length))
dmv$strategy <- as.factor(dmv$strategy)
```

@@

```{r check-n}
length(unique(dm$p_holder))
nrow(dmv)
# dmv <- na.omit(dmv) # some permit holders are lacking vessel length 
# nrow(dmv)
```

Note that we are losing the most permit holders by needing to have data for vessel length.

We'll fit a model with `lmer()`.

```{r fit-model}
# m3 <- lmer(log(cv) ~ scaled_log_spec + scaled_log_days + scale(I(log_specDiv^2)) +
#     (1 + scaled_log_spec+scale(I(log_specDiv^2))|strategy), data = dmv)

# m3 <- lmer(log(cv) ~ scaled_log_spec + scaled_log_days + scaled_log_length +
#     (1 + (scaled_log_spec + scaled_log_days + scaled_log_length)|strategy), data = dmv)

    # scale(I(log_specDiv^2)) +

# m4 <- lmer(log(cv) ~ (scaled_log_spec+scaled_log_days+scaled_log_length)^2 + 
#     (1 + (scaled_log_spec+scaled_log_days+scaled_log_length)^2|strategy), data = dmv)

m4 <- lmer(log(cv) ~ (scaled_log_spec+scaled_log_days)^2 + 
    (1 + (scaled_log_spec+scaled_log_days)^2|strategy), data = dmv)
```

```{r extract-coefficients}
co <- coef(m4)$strategy
co$strategy <- row.names(co)
# names(co)[1:4] <- c("intercept", "log_diversity", "log_days", "strategy")
# arm::display(m4)
# summary(m4)
```

Let's look at the residuals in a few ways.

```{r check-residuals-groups, fig.width=12, fig.height=9}
dmv$residuals <- residuals(m4)
dmv %>%
  arrange(strategy, log_specDiv) %>%
  ggplot(aes(scaled_log_spec, residuals, color = scaled_log_days)) + facet_wrap(~strategy) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE, color = "grey50")
```

These look pretty good with the exception that we might be missing a quadratic effect of species diversity for the strategies involving salmon (`S`). See the curve downward towards the right in the panels that include salmon, especially the salmon-only panel.

```{r check-residuals}
dmv %>%
  arrange(strategy, log_specDiv) %>%
  ggplot(aes(scaled_log_spec, residuals, color = scaled_log_days)) + 
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE, color = "grey50")
```

```{r check-residuals-fitted, fig.width=12, fig.height=9}
dmv$fitted <- fitted(m4)
dmv %>%
  ggplot(aes(fitted, residuals, color = scaled_log_spec)) + facet_wrap(~strategy) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE, color = "grey50")
```

These look pretty good with the exception of the `G S` (herring-salmon) and `S` (salmon-only) where we are under predicting CV in some cases where there is actually high CV.

Across permit holder effects:

```{r coefficient-plot}
broom::tidy(m4, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  filter(!grepl("cor_", term)) %>%
  mutate(conf.low = ifelse(is.na(std.error), estimate, conf.low)) %>%
  mutate(conf.high = ifelse(is.na(std.error), estimate, conf.high)) %>%
  ggplot(aes(y = estimate, ymax = conf.low, ymin = conf.high, x = term)) +
  geom_pointrange() +
  coord_flip() +
  geom_hline(yintercept = 0, lty = 2)
```

What I see:

- increased diversity = lower variability (somewhat quadratically for those including salmon in their strategy)
- longer fishing seasons = lower variability
- benefit of diversifying is much greater for those with longer fishing seasons

Across strategy effects:

```{r group-level-predictors, fig.width=10, fig.height=8}
co <- mutate(co, salmon = grepl("S", strategy))

co_strategy_long <- tidyr::gather(co, parameter, estimate, -salmon, -strategy) %>%
  inner_join(select(dmv, strategy, mean_log_spec) %>% unique())

ggplot(co_strategy_long, aes(mean_log_spec, estimate, label = strategy, color = salmon)) +
  geom_text(size = 3) +
  facet_wrap(~parameter, scales = "free_x") +
  geom_hline(yintercept = 0, lty = 2) +
  geom_smooth(method = "lm", se = FALSE)
```

What I see:

- strategies with higher mean diversity may have very slightly higher CV (but see below where that isn't as clear across most strategy groups)
- strategies with higher mean diversity have more benefit to diversifying if salmon is included in the strategy (this remains consistent across the groups in the below figure)

```{r group-level-predictors-grouped, fig.width=14, fig.height=7}
strategy_grouped <- rbind(cbind(co_strategy_long[grep("M",co_strategy_long$strategy),], group="Groundfish"),
  cbind(co_strategy_long[grep("S",co_strategy_long$strategy),], group="Salmon"),
  cbind(co_strategy_long[grep("B|C",co_strategy_long$strategy),], group="Halibut-Sablefish"),
  cbind(co_strategy_long[grep("K|T|D",co_strategy_long$strategy),], group="Crab"),
  cbind(co_strategy_long[grep("G",co_strategy_long$strategy),], group="Herring"))

strategy_grouped <- filter(strategy_grouped, 
  parameter %in% c("(Intercept)", "scaled_log_days", "scaled_log_spec"))
ggplot(strategy_grouped, aes(mean_log_spec, estimate, label = strategy)) +
  geom_text() + geom_smooth(se = FALSE, method = "lm") +
  geom_hline(yintercept = 0, lty = 2) +
  facet_grid(parameter~group, scales = "free_y")

```
